\chapter{The scale of meditation of an author, ancient but nameless.}
\label{appendix:scale}
\begin{center}\emph{Degree of Preparation.}\end{center} \par
    \begin{enumerate}
        \item \begin{description}\item[Question.] What I $ \begin{cases} \text{think.} \\ \text{should think.} \end{cases} $ \end{description}
        \item \begin{description}\item[Excussion.] A repelling of what I should not think. \end{description}
%%        \item \begin{description}[style=multiline]\item[Choice \\ or \\ Election.] $\Bigg\}$ Of what most $ \begin{cases} \text{necessary.} \\ \text{expedient.} \\ \text{comely.} \end{cases} $\end{description}
        \item \parbox[c]{1.5cm}{\textbf{Choice\\or\\Election.}} \parbox[c]{10.5cm}{ $\Bigg\}$ Of what most $ \begin{cases} \text{necessary.} \\ \text{expedient.} \\ \text{comely.} \end{cases} $}
    \end{enumerate}

    \begin{center}\emph{Degrees of Proceeding in the Understanding.}\end{center} \par
    \begin{enumerate}\addtocounter{enumi}{3}
        \item \begin{description}\item[Commemoration] An actual thinking upon the matter elected.\end{description}
        \item \begin{description}\item[Consideration] A redoubled commemoration of the same till it be fully known.\end{description}
        \item \begin{description}\item[Attention] A fixed and earnest consideration, whereby it is fastened in the mind.\end{description}
        \item \begin{description}\item[Explanation] A clearing of the thing considered by similitudes.\end{description}
        \item \begin{description}\item[Tractation] An extending the thing considered to other points, where all questions of doubts are discussed.\end{description}
        \item \begin{description}\item[Dijudication] An estimation of the worth of the thing thus handled.\end{description}
        \item \begin{description}\item[Causation] A confirmation of the estimation thus made.\end{description}
        \item \begin{description}\item[Rumination] A sad and serious meditation of all the former, till it may work upon the affections.\end{description}        
    \end{enumerate}
\begin{center}From hence to the degrees of affection.\end{center}
