\chapter{The practice of meditation.}
\section{Wherein, first, we begin with some description of that we meditate of.}
First, therefore, it shall be expedient to consider seriously, what the thing is whereof we meditate. 

``What then, my soul, is the life of the saints, whereof thou studiest? Who are the saints, but those which, having been weakly holy upon earth, are perfectly holy above? Which even on earth were perfectly holy in their Saviour, now are so in themselves? Which, overcoming on earth, are truly canonized in heaven? What is their life, but that blessed estate above, wherein their glorified soul hath a full fruition of God?'' 

\section{Secondly, follows an easy and voluntary division of the matter meditated.}

The nature whereof, after we have thus shadowed out to ourselves by a description; not curious always, and exactly framed according to the rules of art, but sufficient for our own conceit; the next is, if it shall seem needful, or if the matter will bear or offer it, some easy and voluntary division, whereby our thoughts shall have more room made for them, and our proceeding shall be more distinct. 

``There is a life of nature; when thou, my soul, dwellest in this body, and informest thine earthly burden: there is a life of grace; when the Spirit of God dwells in thee: there is a life of glory; when the body being united to thee, both shall be united to God; or when, in the mean time, being separated from thy companion, thou enjoyest God alone. This life of thine therefore, as the other hath his ages, hath his statures: for it entereth upon his birth when thou passest out of thy body, and changest this earthly house for an heavenly: it enters into his full vigour, when at the day of the common resurrection, thou resumest this thy companion; unlike to itself, like to thee, like to thy Sayiour; immortal now, and glorious. In this life here, may be degrees; there, can be no imperfection. If some be like the sky, others like the stars; yet all shine. If some sit at their Saviour's right hand, others at his left; all are blessed. If some vessels hold more, all are full; none complaineth of want, none envieth him that hath more.''

\section{A consideration of the causes thereof, in all kinds of them.}
Which done, it shall be requisite for our perfecter understanding, and for the laying grounds of matter for our affection, to carry it through those other principal places and heads of reason which nature hath taught every man, both for knowledge and amplification; the first whereof are the causes, of all sorts. 

``Whence is this eternal life, but from him which only is eternal; which only is the fountain of life; yea, life itself? Who but the same God that gives our temporal life giveth also that eternal? The Father bestoweth it, the Son meriteth it, the Holy Ghost seals and applieth it. Expect it only from him, my soul, whose free election gave thee the first title to it, to be purchased by the blood of thy Saviour. For thou shalt not therefore be happy, because he saw that thou wouldest be good; but therefore art thou good, because he hath ordained thou shalt be happy. He hath ordained thee to life; he hath given thee a Saviour, to give this life unto thee; faith, whereby thou mightest attain to this Saviour; his word, by which thou mightest attain to this faith: what is there in this not his? And yet, not his so simply, as that it is without thee; without thy merit indeed, not without thine act. Thou livest here, through his blessing, but by bread; thou shalt live above, through his mercy, but by thy faith below, apprehending the Author of thy life. And yet, as he will not save thee without thy faith, so thou canst never have faith without his gift. Look to him, therefore, O my soul, as the beginner and finisher of thy salvation; and while thou magnifiest the Author be ravished with the glory of the work, which far passeth both the tongue of angels and the heart of man. It can no be good thing that is not there. How can they want water that have the spring? Where God is enjoyed, in whom only all things are good, what good can be wanting. And what perfection of bliss is there where all goodness is met and united! \emph{In thy presence is fulness of joy; and at thy right hand are pleasures for evermore.} O blessed reflection of glory! We see there as we are seen; in that we are seen, it is our glory; in that we see, it is God's glory; therefore doth he glorify us, that our glory should be to his. How worthy art thou, Lord, that through us thou shouldest look at thyself!''

\section{The consideration of the fruits and effects.}

The next place shall be the fruits and effects following upon their several causes; which also affords very feeling and copious matter to our meditation; wherein it shall be ever best not so much to seek for all, as to choose out the chiefest. 

``No marvel then, if from this glory proceed unspeakable joy; and from this joy, the sweet songs of praise and thanksgiving. The Spirit bids us, when we are merry, sing; how much more then, when we are merry without all mixture of sorrow, beyond all measure of our earthly affections, shall we sing joyful hallelujahs and hosannahs to him that dwelleth in the highest heavens! Our hearts should be so full that we cannot choose but sing, and we cannot but sing melodiously. There is no jar in this music, no end of this song. O blessed change of the saints! They do nothing but weep below, and now nothing but sing above. We sowed in tears, reap in joy; there was some comfort in those tears when they were at worst, but there is no danger of complaint in this heavenly mirth. If we cannot sing here with angels, \emph{On earth peace,} yet there we shall sing with them, \emph{Glory to God on high;} and, joining our voices to theirs, shall make up that celestial concert which none can either hear or bear part in and not be happy.''

\section{Consideration of the subject wherein or whereabout it is.}

After which comes to be considered the subject, either wherein that is, or whereabout that is employed, which we meditate of; as, 

``And indeed what less happiness doth the yery place promise wherein this glory is exhibited? which is no other than the paradise of God. Here below we dwell, or rather we wander, in a continued wilderness; there we shall rest us in the true Eden; \emph{I am come into my garden, my sister, my spouse.} Kings use not to dwell in cottages of clay, but in royal courts fit for their estate; how much more shall the King of heaven, who hath prepared for men so fair mansions on earth, make himself an habitation suitable to his Majesty! Even earthly princes have dwelt in cedar and ivory; but the great city, Holy Jerusalem, the palace of the Highest, hath her walls of jasper, her building of gold, her foundation of precious stones, her gates of pearl: \emph{How glorious things are spoken of thee, O thou city of God!} We see but the pavement, and yet how goodly it is! The believing Centurion thought himself unworthy that Christ should come under his roof; yet wert thou, Saviour, in thine humbled estate, in the form of a servant; how then shall I think myself worthy to come under this roof of thine, so shining and glorious? If this clay of mine may come to this honour above, let it be trampled upon and despised on earth''

\section{Consideration of the appendances and qualities of it.}

Sixthly shall follow the appendances and qualities which cleave unto the subject whereof we meditate; as, 

``But were the place less noble and majestical, yet the company which it affordeth hath enough to make the soul blessed; for not the place giveth ornament to the guest so much as the guest to the place. How loath are we to leave this earth only for the society of some few friends in whom we delight, which yet are subject every day to mutual dislikes! what pleasure shall we then take in the enjoying of the saints, when there is nothing in them not amiable, nothing in us that may cool the fervour of our love! There shalt thou, my soul, thyself glorified, meet with thy dear parents and friends alike glorious, never to be severed. There thou shalt see and converse with those ancient worthies of the former world, the blessed patriarchs and prophets, with the crowned martyrs and confessors, with the holy apostles and the fathers of that primitive and this present church, shining each one according to the measure of his blessed labours. There shalt thou live farmiliarly in the sight of those angels whom now then receivest good from, but seest not. There, which is the head of all thy felicity, thine eyes shall see Him whom now thy heart longeth for; that Saviour of thine, in the only hope of whom now thou livest. Alas! How dimly and afar off dost thou now behold him! How imperfectly dost thou enjoy him, while every temptation bereaves thee for the time of his presence! \emph{I sought him whom my soul loveth; I sought him, but found him not.} His back is now towards thee many times through thy sins; and therefore thou hardly discernest him. Otherwhile, and often, thy back is turned imto him through negligence, that when thou mightest obscurely see him thou dost not; now thou shalt see him, and thine eyes, thus fixed, shall not be removed. Yet neither could this glory make us happy, if, being thus absolute, it were not perpetual. To be happy is not so sweet a state, as it is miserable to have been happy. Lest aught therefore should be wanting, behold, this felicity knoweth no end, feareth no intermission, and is as eternal for the continuance as he that had no beginning. O blessedness truly infinite! Our earthly joys do scarce ever begin, but when they begin, their end bordereth upon their beginning. One hour seeth us ofttimes joyful and miserable; here alone is nothing but eternity. If then the divine prophet thought here one day in God's earthly house better than a thousand otherwhere, what shall I compare to thousands of millions of years in God's heavenly temple? Yea, millions of years are not so much as a minute to eternity, and that other house not a cottage to this.''

\section{Of that which is diverse from it, or contrary to it.}

Seventhly, our thoughts, leaving a while the consideration of the thing as it is in itself, shall descend unto it as respectively with others; and therefore, first, shall meditate of that which is diverse from it, or contrary unto it. 

``What dost thou here then, my soul? what dost thou here grovelling upon earth, where the best things are vanity, the rest no better than vexation? Look round about thee, and see whether thine eyes can meet with any thing but either sins or miseries. Those few and short pleasures thou seest end ever sorrowfully, and in the mean time are intermingled with many grievances. Here thou hearest one cry out of a sick body, whereof there is no part which affords not choioe of diseases; this man layeth his hand upon his consuming lungs, and complaineth of short wind; that other, upon his rising spleen; a third shaketh his painful head; another roars out for the torment of his reins or bladder; another, for the racking of his gouty joints: one is distempered with a watery dropsy; another, with a windy colic; a third, with a fiery ague; a fourth, with an earthen melancholy: one grovels and foameth with the falling sickness; another lieth bedrid, half senseless, with a dead palsy: there are but few bodies that complain not of some disease; and, that thou mayest not look far, it is a wonder if thyself feel not always one of these enh within thee. There thou hearest another lament his loss; either his estate is impaired by suretyship, or stealth, or shipwreck, or oppression; or his child is unruly, or miscarried; or his wife dead or disloyal: another tormented with passions: each one is some way miserable. But that which is yet more irksome, thy one ear is beaten with cursings and blasphemies; thy other with scornful, or wanton, or murdering speeches; thine eyes see nothing but pride, filthiness, profaneness, blood, excess, and whatsoever else might vex a righteous soul; and, if all the world besides were innocent, thou findest enough within thyself to make thyself weary, and thy life loathsome. Thou needest not fetch cause of complaint from others: thy corruptions yield thee too much at home; ever sinning, ever presuming: sinning, even when thou hast repented, yea, even while thou repentest, sinning. Go to now, my soul, and solace thyself here below; and suffer thyself besotted with these goodly contentments, worthy of no better, while thou fixest thyself on these. See if thou canst find any of these above; and, if thou canst meet with any distemper, any loss, any sin, any complaint, from thyself or any other above, despise thine heaven as much as now thou lovest the earth. Or, it all this cannot enough commend unto thee the state of heavenly glory, cast down thine eyes yet lower, into that deep and bottomless pit, full of horror, full of torment: where there is nothing but flames, and tears, and shrieks, and gnashing of teeth; nothing but fiends and tortures: where there is palpable darkness, and yet perpetual fire; where the damned are ever boiling, never consumed; ever dying, never dead; ever complaining, never pitied: where the glutton, that once would not give a crust of bread, now begs for one drop of water; and yet, alas! If whole rivers of water should fall into his mouth, how should they quench those rivers of brimstone that feed this flame! Where there is no intermission of complaints; no breathing from pain; and, after millions of years, no possibility of comfort. And if the rod wherewith thou chastisest thy children, O Lord, even in this life, be so smart and galling, that they have been brought down to the brim of despair, and in the bitterness of their soul have entreated death to release them; what shall I think of their plagues, in whose righteous confusion, thou insultests\footnote{The word here replaced from the editions of 1614 and 1628 is so offensive to our ears, from the idea which its present use associates with it, that one would gladly substitute for it a less objectionable equivalent, but that equivalent is not to be found in ``consultest,'' which is the reading of later editions. The word is clearly intended to represent the righteous vengeance of the Almighty triumphing over his enemies.}, and sayest, \emph{Aha, I will avenge me of mine enemies?} Even that thou shalt not be thus miserable, O my soul, is some kind of happiness; but that thou shalt be as happy as the reprobate are miserable, how worthy is it of more estimation than thyself is capable of!''

\section{Of comparisons and similitudes, whereby it may be most fitly set forth.}

After this opposition, the mind shall make comparison of the matter meditated with what may nearest resemble it, and shall illustrate it with fittest similitudes, which give no small light to the understanding, nor less force to the affection. 

``Wonder then, O my soul, as much as thou canst, at this glory; and in comparison thereof contemn this earth, which now thou treadest upon; whose joys, if they were perfect, are but short; and if they were long, are imperfect. One day, when thou art above, looking down from the height of thy glory, and seeing the sons of men creeping like so many ants on this molehill of earth, thou shalt think, `Alas, how basely I once lived I was yonder silly dungeon the place I so loved, and was so loath to leave?' Think so now beforehand; and, since of heaven thou canst not, yet account of the earth as it is worthy: how heartless and irksome are ye, ye best earthly pleasures, if ye be matched with the least of those above! How vile are you, ye sumptuous buildings of kings, even if all the entrails of the earth had agreed to enrich you, in comparison of this frame not made with hands! It is not so high above the earth, in distance of place, as in worth and majesty. We may see the face of heaven from the heart of the earth; but from the nearest part of the earth, who can see the least glory of heaven? The three disciples on mount Tabor saw but a glimpse of this glory shining upon the face of their Saviour; and yet, being ravished with the sight, cried out, \emph{Master, it is good being here;} and, thinking of building of three tabernacles, (for Christ, Moses, Elias,) could have been content themselves to have lien without shelter, so they might always have enjoyed that sight. Alas I how could earthly tabernacles have fitted those heavenly bodies? They knew what they saw: what they said, they knew not. Lo, these three disciples were not transfigured  yet how deeply they were affected even with the glory of others! How happy shall we be, when oarselves shall be changed into glorious; and shall have tabernacles, not of our own making, but prepared for us by God! And yet not tabernacles, but eternal mansions: Moses saw God but a while, and shined: how shall we shine, that shall behold his face forever? What greater honour is there than in sovereignty? What greater pleasure than in feasting? This life is both a kingdom and a feast! A kingdom: \emph{He that overcomes shall rule the nations; and shall sit with me in my throne:} O blessed promotion! O large dominion and royal seat! To which Solomon's throne of ivory was not worthy to become a footstool. A feast: \emph{Blessed are they that are called to the marriage-supper of the Lamb:} feasts have more than necessity of provision, more than ordinary diet; but marriage-feasts yet more than common abundance; but the marriage-feast of the Son of God to his blessed spouse, the Church, must so far exceed in all heavenly munificence and variety, as the persons are of the greater state and majesty: there is new wine, pure manna, and all manner of spiritual dainties; and, with the continual cheer, a sweet and answerable welcome; while the Bridegroom lovingly cheereth us up, \emph{Eat, friends; drink, and make you merry O well beloved:} yea, there shalt thou be, my soul, not a guest, but, how unworthy soever, the bride herself, whom he hath everlastingly espoused to himself in truth and righteousness. The contract is passed here below; the marriage is consummate above, and solemnized with a perpetual feast: so that now thou mayest safely say, \emph{My well-beloved is mine, and I am his: wherefore hearken, O my soul, and consider, and incline thine ear; forget also thine own people, and thy father's house,} thy supposed home of this world: so shall the King have pleasure in thy beauty; for he is the Lord, and worship thou him.'' 

\section{The titles and names of the thing considered.}

The very names and titles of the matter considered yield no small store to our meditation: which being commonly so imposed, that they secretly comprehend the nature of the thing which they represent, are not unworthy of our discourse. 

``What need I seek those resemblances, when the very name of life implieth sweetness to men on earth, even to them which confess to live with some discontentment? \emph{Surely the light is a pleasant thing; and it is good to the eyes to see the sun?:} yet when temporal is added to life, I know not how, this addition detracteth something, and doth greatly abate the pleasure of life; for those which joy to think of life, grieve to think it but temporal: so vexing is the end of that whose continuance was delightful. But now, when there is an addition, above time, of eternity, it maketh life so much more sweet as it is more lasting: and, lasting infinitely, what can it give less than an infinite contentment? O dying and false life which we enjoy here, and scarce a shadow and counterfeit of that other! What is, more esteemed than glory? Which is so precious to men of spirit, that it makes them prodigal of their blood, proud of their wounds, careless of themselves: and yet, alas! howfpent and how fading is this glory, effected with such dangers and death; hardly, after all trophies and monuments, either known to the next sea, or surviving him that dieth for it! It is true glory to triumph in heaven, where is neither envy, nor forgetfulness. What is more dear to us than our country? Which the worthy and faithful patriots of all times have respected above their parents, their children, their lives; counting it only happy to live in it, and to die for it: the banished man pines for the want of it: the traveller digesteth all the tediousness of his way, all the sorrows of an ill journey, in the only hope of home, forgetting all his foreign miseries when he feeleth his own smoke. Where is our country, but above? Thence thou camest, O my soul; thither thou art going, in a short but weary pilgrimage. O miserable men, if we account ourselves at home in our pilgrimage, if in our journey we long not for home! Dost thou see men so in love with their native soil, that, even when it is all deformed with the desolations of war and turned into rude heaps, or while it is even now flaming with the fire of civil broils, they covet yet still to live in it, preferring it to all other places of more peace and pleasure? and shalt thou, seeing nothing but peace and blessedness at home, nothing but trouble abroad, content thyself with a faint wish of thy dissolution? If heaven were thy gaol, thou couldest but think of it uncomfortably. O what affection can be worthy of such an home!''

\section{Consideration of fit testimonies of Scripture concerning our theme.}
Lastly, if we can recall any pregnant testimonies of Scripture concerning our theme, those shall fitly conclude this part of our meditation: of Scripture; for that in these matters of God none but divine authority can command assent, and settle the conscience. Witnesses of holy men may serve for colours, but the ground must be only from God. 

``\emph{There it is,} saith the Spirit of God which cannot deceive thee, \emph{that all tears shall be wiped from our eyes; there shall be no more death, nor sorrow, nor crying, neither shall there be any more pain;} yea, there shall not only be an end of sorrows, but an abundant recompense for the sorrows of our life; as he that was rapt up into the third heaven, and there saw what cannot be spoken, speaketh yet thus of what he saw: \emph{I count, that the afflictions of this present time are not worthy of the glory which shall be showed to us.} It was showed unto him what should hereafter be showed unto us; and he saw that, if all the world full of miseries were laid in one balance, and the least glory of heaven in another, those would be incomparably light; yea, as that divine father, that one day's felicity above were worth a thousand years' torment below. What then can be matched with the eternity of such joys? \emph{O how great,} therefore, \emph{is this thy goodness, Lord, which thou hast laid up for them, that fear thee;} and done to them that trust in thee, before the sons of men!''
