\chapter{The conclusion of our meditation, in what order it must be.}
\section{First, with thanksgiving.}

The conclusion remaineth: wherein we must advise, like as physicians do in their sweats and exercise, that we cease not over-suddenly, but leave off by little and little. The mind may not be suffered to fall headlong from this height, but must also descend by degrees. 

The first whereof, after our confidence, shall be an hearty gratulation and thanksgiving; for as man naturally cannot be miserable, but he must complain and crave remedy; so the good heart cannot find itself happy and not be thankful; and this thankfulness, which it feeleth and expresseth, maketh it yet more good, and affecteth it more. 

``What shall I then do to thee for this mercy, O thou Saviour of men? What should I render to my Lord for all his benefits? Alas! What can I give thee which is not thine own before? O that I could give thee but all thine! Thou givest me to drink of this cup of salvation; \emph{I will, therefore, take the cup of salvation, and call upon the name of the Lord\footnote{Psalms 116:13 ``I will take the cup of salvation, and call upon the name of the LORD.''}: praise thou the Lord, O my soul; and all that is within me praise his holy name\footnote{Psalms 103:1 ``(A Psalm of David.) Bless the LORD, O my soul: and all that is within me, bless his holy name.'' }.} And since here thou beginnest thy heaven, begin here also that joyful song of thanksgiving which there thou shalt sing more sweetly, and never end.''

\section{Secondly, with recommendation of our souls and ways to God.}

After this thanksgiving shall follow a faithful recommendation of ourselves to God; wherein the soul doth cheerfully give up itself, and repose itself wholly upon her Maker and Redeemer; committing herself to him in all her ways; submitting herself to him in all his ways; desiring in all things to glorify him, and to walk worthy of her high and glorious calling. 

Both which latter shall be done, as I have ever found, with much life and comfort, if, for the full conclusion we shall lift up our heart and voice to God, in singing some versicle of David's divine Psalms, answerable to our disposition and matter; whereby the heart closes up itself with much sweetness and contentment. 

This course of meditation thus heartily observed, let him that practiseth it tell me whether he find not that his soul, which, at the beginning of this exercise did but creep and grovel upon earth, do not now, in the conclusion, soar aloft in heaven, and, being before aloof off, do not now find itself near to God, yea with him and in him. 