\chapter{The benefit and uses of meditation.}
\section{Which are universal to all Christians, and not to be appropriated to some professions.}

It is not, I suppose, a more bold than profitable labour, after the endeavours of so many contemplative men, to teach the art of meditation: an heavenly business as any that belongeth either to man or Christian; and such as whereby the soul doth unspeakably benefit itself. For by this do we ransack our deep and false hearts; find out our secret enemies; buckle with them, expel them; arm ourselves against their re-entrance: by this, we make use of all good means, fit ourselves to all good duties; by this we descry our weakness; obtain redress; prevent temptations; cheer up our solitariness; temper our occasions of delight; get more light unto our knowledge, more heat to our affections, more life to our devotion: by this, we grow to be, as we are, strangers upon earth; and out of a right estimation of all earthly things into a sweet fruition of invisible comforts: by this, we see our Saviour, with Stephen; we talk with God, as Moses: and by this we are ravished, with blessed Paul, into paradise; and see that heaven, which we are loath to leave, which we cannot utter. This alone is the remedy of security and worldliness, the pastime of saints, the ladder of heaven, and, in short, the best improvement of Christianity. Learn it who can, and neglect it who list; he shall never find joy, neither in God nor in himself, which doth not both know and practise it. 

And however of old some hidden cloisters have engrossed it to themselves and confined it within their cells, who indeed, professing nothing but contemplation, through their immunity from those cares which accompany an active life, might have the best leisure to this business; yet, seeing there is no man so taken up with action as not sometimes to have a free mind; and there is no reasonable mind so simple as not to be able both to discourse somewhat and to better itself by her secret thoughts; I deem it an envious wrong to conceal that from any whose benefit may be universal. Those that have but a little stock had need to know the best rules of thrift. 

