\newleadpage{meditation}{A Meditation of Death}
\meditation
\chapter{An Example Meditation, According To The Former Rules}
\section{The entrance.}
And now, my soul, that thou hast thought of the end, what can fit thee better than to think of the way? And though the forepart of the way to heaven be a good life, the latter and more immediate is death. Shall I call it the way, or the gate of life? Sure I am, that by it only we pass into that blessedness; whereof we have so thought, that we have found it cannot be thought of enough. 

\section{The description.}
What then is this death but the taking down of these sticks, whereof this earthly tent is composed? the separation of two great and old friends, till they meet again? the gaol-delivery of a long prisoner? our journey into that other worlds for which we and this thoroughfare were made? our payment of our first debt to nature; the sleep of the body and the awaking of the soul? 

\section{The division}
But, lest thou shouldest seem to flatter him whose name and face hath ever seemed terrible to others, remember that there are more deaths than one: if the first death be not so fearful as he is made, his horror lying more in the conceit of the beholder than in his own aspect, surely the second is not made so fearful as he is. No living eye can behold the terrors thereof; it is as impossible to see them, as to feel them and live. Nothing but a name is common to both. The first hath men, casualties, diseases, for his executioners; the second, devils: the power of the first is in the grave; the second, in hell: the worst of the first is senselessness; the easiest of the second is a perpetual sense of all the pain that can make a man exquisitely miserable. 

\section{The causes.}
Thou shalt have no business, O my soul, with the second death: thy first resurrection hath secured thee. Thank him that hath redeemed thee for, thy safety. And how can I thank thee enough, O my Saviour, which hast so mercifully bought off my torment with thy own ; and hast drunk off that bitter potion of thy Father's wrath, whereof the very taste had been our death? Yea, such is thy mercy, O thou Redeemer of men, that thou hast not only subdued the second death, but reconciled the first: so as thy children taste not at all of the second; and find the first so sweetened to them by thee, that they complain not of bitterness. It was not thou, O God, that madest death: our hands are they that were guilty of this evil. Thou sawest all thy work that it was good: we brought forth sin, and sin brought forth death. To the discharge of thy justice and mercy we acknowledge this miserable conception: and needs must that child be ugly that hath such parents. Certainly, if being and good be, as they are, of an equal extent, then the dissolution of our being must needs in itself be evil. How full of darkness and horror then is the privation of this vital light, especially since thy wisdom intended it to the revenge of sin, which is no less than the violation of an infinite justice! It was thy just pleasure to plague us with this brood of our own begetting. Behold, that death, which was not till then in the world, is now in every thing: one great conqueror finds it in a slate; another finds it in a fly: one finds it in the kernel of a grape; another in the prick of a thorn: one, in the taste of an herb; another, in the smell of a flower: one, in a bit of meat; another, in a mouthful of air: one, in the very sight of a danger; another, in the conceit of what might have been. Nothing in all our life is too little to hide death under it. There need no cords, nor knives, nor swords, nor pieces: we have made ourselves as many ways to death as there are helps of living. But if we were the authors of our death, it was thou that didst alter it: our disobedience made it; and thy mercy made it not to be evil. It had been all one to thee to have taken away the very being of death from thine own; but thou thoughtest it best to take away the sting of it only: as good physicians, when they would apply their leeches, scour them with salt and nettles; and when their corrupt blood is voided, employ them to the health of the patient. It is more glory to thee that thou hast removed enmity from this Esau; that now he meets us with kisses instead of frowns: and if we receive a blow from this rough hand, yet that very stripe is healing. O how much more powerful is thy death than our sin! O my Saviour, how hast thou perfumed and softened this bed of my grave by dying! How can it grieve me to tread in thy steps to glory? 

\section{The effects.}
Our sin made death our last enemy: thy goodness hath made it the first friend that we meet with in our passage to another world: for, as she that receives us from the knees of our mother in our first entrance to the light, washeth, cleanseth, dresseth us, and presents us to the breast of our nurse or the arms of our mother, challenges some interest in us when we come to our growth; so death, which, in our passage to that other life is the first that receives and presents our naked souls to the hands of those angels which carry it up to her glory, cannot but think this office friendly and meritorious. What, if this guide lead my carcass through corruption and rottenness, when my soul, in the very instant of her separation knows itself happy? What if my friends mourn about my bed and coffin, when my soul sees the smiling face and loving embracements of him that was dead and is alive? What care I who shuts these earthen eyes when death opens the eye of my soul to see as I am seen? What if my name be forgotten of men, when I live above with the God of spirits? 

\section{The subject.}
If death would be still an enemy, it is the worst part of me that he hath any thing to do withal; the best is above his reach; and gains more than the other can lose. The worst piece of the horror of death is the grave: and, set aside infidelity, what so great misery is this? That part which is corrupted feels it not: that which is free from corruption feels an abundant recompense, and foresees a joyful reparation. What is here, but a just restitution? We carry heaven and earth wrapt up in our bosoms; each part returns homeward; and if the exceeding glory of heaven cannot countervail the dolesomeness of the grave, what do I believing? But, if the beauty of that celestial sanctuary do more than equalize the horror of the bottomless pit, how can I shrink at earth like myself, when I know my glory? And if examples can move thee any whit, look behind thee, O my soul, and see which of the worthies of that ancient latter world, which of the patriarchs, kings, prophets, apostles, have not trod in these red steps. Where are those millions of generations which have hitherto peopled the earth? How many passing-bells hast thou heard for thy known friends? How many sick beds hast thou visited? How many eyes hast thou seen closed? How many vain men hast thou seen that have gone into the field to seek death, in hope to find an honour as foolish as themselves? How many poor creatures hast thou mulcted\footnote{Mulct --- to punish by a fine} with death for thine own pleasure? And canst thou hope that God will make a by-way and a postern for thee alone, that thou mayest pass to the next world, not by the gates of death, not by the bottom of the grave? 
 
\section{The adjunct.}
What then dost thou fear, O my soul? There are but two stages of death, the bed and the grave; this latter, if it have senselessness, yet it hath rest; the former, if it have pain, yet it hath speediness; and when it lights upon a faithful heart meets with many and strong antidotes of comfort. The evil that is ever in motion, is not fearful: that which both time and eternity finds standing where it was is worthy of terror. Well may those tremble at death which find more distress within than without; whose consciences are more sick and nearer to death than their bodies. It was thy Father's wrath that did so terrify thy soul, O my Saviour, that it put thy body into a bloody sweat\footnote{Luke 22:44 ``And being in an agony he prayed more earnestly: and his sweat was as it were great drops of blood falling down to the ground.''}. The mention and thought of thy death ended in a psalm; but this began in an agony. Then didst thou sweat out my fears. The power of that agony doth more comfort all thine, than the angels could comfort thee. That very voice deserved an eternal separation of horror from death, where thou saidst, \emph{My God, my God, why hast thou forsaken me?\footnote{Psalms 22:1 ``(To the chief Musician upon Aijeleth Shahar, A Psalm of David.) My God, my God, why hast thou forsaken me? why art thou so far from helping me, and from the words of my roaring?''}} Thou hadst not complained of being left, if thou wouldest have any of thine left destitute of comfort in their parting. I know not whom I can fear while I know whom I have believed: how can I be discouraged with the sight of my loss, when I see so clear an advantage? 

\section{The contrary.}
What discomfort is this, to leave a frail body to be joined unto a glorious head? To forsake vain pleasures, false honours, bootless hopes, unsatisfying wealth, stormy contentments, sinful men, perilous temptations, a sea of troubles, a galley of servitude, an evil world and a consuming life, for freedom, rest, happiness, eternity? And if thou wert sentenced, O my soul, to live a thousand years in this body with these infirmities, how wouldst thou be weary; not of being only, but of complaining: where, ere the first hundred, I should be a child; ere the second, a beast; a stone, ere the third; and therefore should be so far from finding pleasure in my continuance, that I should not have sense enough left to feel myself miserable! And when I am once gone, what difference is there betwixt the agedest of the first patriarchs and me, and the child that did but live to be born, save only in what was! And that which was is not. And if this body had no weakness to make my life tedious, yet, what a torment is it, that while I live I must sin! Alas, my soul, every one of thy known sins is not a disease, but a death! What an enemy art thou to thyself, if thou canst not be content, that one bodily death should excuse thee from many spiritual; to cast off thy body that thou mayest be stripped of the rags, yea, the fetters of thy sin, and clothed with the robes of glory? Yet these terms are too hard: thou shalt not be cast off, O my body: rather thou shalt be put to making. This change is no less happy for thee than for thy partner. This very skin of thine, which is now tawny and wrinkled, shall once shine: this earth shall be heaven: this dust shall be glorious: these eyes, that are now weary of being witnesses of thy sins and miseries, shall then never be weary of seeing the beauty of thy Saviour, and thine own in his: these ears, that have been now tormented with the impious tongues of men, shall first hear the voice of the Son of God; and then the voices of saints and angels in their songs of hallelujah: and this tongue, that now complains of miseries and fears, shall then bear a part in that divine harmony.

\section{The comparisons.}
In the mean time thou shalt but sleep in this bed of earth. He that hath tried the worst of death hath called it no worse. Very heathens have termed them cousins; and it is no unusual thing for cousins of blood to carry both the same names and features. Hast thou wont, O my body, when the day hath wearied thee, to he down not unwillingly to thy rest? behold, in this sleep there is more quietness, more pleasure of visions, more certainty of waking, more cheerfulness in rising; why then art thou loath to think of laying off thy rags and reposing thyself? Why art thou like a child unwilling to go to bed? Hast thou ever seen any bird, which, when the cage hath been opened, would rather sit still and sing within her grates, than fly forth unto her freedom in the woods? Hast thou ever seen any prisoner in love with his bolts and fetters? Did the chief of the apostles, when the angel of God shined in his gaol, and struck him on the side, and loosed his two chains, and bade him arise quickly\footnote{Acts 12:7 ``And, behold, the angel of the Lord came upon him, and a light shined in the prison: and he smote Peter on the side, and raised him up, saying, Arise up quickly. And his chains fell off from his hands.''}, and opened both the wooden and iron gate, say, ``What! So soon? Yet a little sleep?'' What madness had it been, rather to slumber betwixt his two keepers, than to follow the angel of God into liberty? Hast thou ever seen any mariner that bath sainted the sea with songs and the haven with tears? What shall I say to this diffidence, O my soul, that thou art unwilling to think of rest after thy toil; of freedom after thy durance; of the haven after an unquiet and tempestuous passage? How many are there that seek death and cannot find it, merely out of the irksomeness of life. Hath it found thee, and offered thee better conditions; not of immunity from evils, but of possession of more good than thou canst think; and wouldest thou now fly from happiness to be rid of it? 

\section{The names.}
What! Is it a name that troubles thee? What if men would call sleep death; wouldest thou be afraid to close thine eyes? What hurt is it then, if he, that sent the first sleep upon man whilst he made him an helper\footnote{Genesis 2:21-22 ``And the LORD God caused a deep sleep to fall upon Adam, and he slept: and he took one of his ribs, and closed up the flesh instead thereof; \textsuperscript{22}And the rib, which the LORD God had taken from man, made he a woman, and brought her unto the man.''}, send this last and soundest sleep upon me while he prepares my soul for a glorious spouse to himself? It is but a parting which we call death; as two friends, when they have led each other on the way, shake hands till they return from their journey. If either could miscarry, there were cause of sorrow; now they are more sure of a meeting than of a parture, what folly is it not to be content to redeem the unspeakable gain of so dear a friend with a little intermission of enjoying him! He will return laden with the riches of heaven; and will fetch his old partner to the participation of this glorious wealth. Go then, my soul, to this sure and gainful traffick; and leave my other half in an harbour as safe, though not so blessed; yet so shalt thou be separated, that my very dust shall be united to thee still, and to my Saviour in thee. 